Software Development in Startup Companies: The Green-field Startup Model

### 2013-02-14 article/main.tex
added validation complete

\author{FirstName~SecondName,~\IEEEmembership{Fellow,~IEEE,}
        BLA~BLA,~\IEEEmembership{Fellow,~IEEE,}
        and~BLA~BLA,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem Author info.\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: see http://something/contact.html
\IEEEcompsocthanksitem Author info.}% <-this % stops a space
\thanks{}}

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
Software startups are newly created companies with no operating history and extremely fast in producing cutting-edge technologies. Despite their increasing importance in economy there are only few scientific studies attempting to address software engineering (SE) issues, especially for early-stage startups. This research aims to understand the software development strategies conducted by practitioners, in the period of time that goes from idea conception to the first open beta release. A \textit{grounded theory} approach guided the execution of a case study to explore the state-of-practice. As result, the Greenfield startup model (GSM)  reveals the urgent priority of startups of releasing the product as quickly as possible to verify the product/market fit and to adjust the business and product trajectory according to early collected user feedbacks. Nevertheless, the need of shortening time-to-market, by speeding-up the development through low-precision and product-centric engineering activities, is counterbalanced by the need of restructuring the product before setting off for further growth.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

Nevertheless, by speeding-up the development process, startups accumulate technical debt, ignoring processes, relying on informal communication and replacing documentation with low-precision artifacts. For startups it is essential to look ahead within short-term deadlines and determine the right product/market fit driven by user feedbacks, using flexible and reactive development approaches and adopting mitigation strategies to contain the severity of accumulated technical debt.

@@

The remain of this paper is structured as follow:  we provide relevant information about startups in section \ref{backg}. Section \ref{resmet} introduces the research questions and discusses the design and execution of the research process. Results are discussed in section \ref{res:gsm} which presents the GSM model. Section \ref{res:val}  compares results of this research to literature and frameworks. Section \ref{valt} discusses the validity threats considered during this study. Finally, this paper concludes in section \ref{conc}.

@@

Looking at the number of new business incubators which appeared in the last three years one can evaluate the importance of startups \cite{Grimaldi2005}. The wave of disruption in new technologies has led companies to be more and more competitive, forcing themselves to radical organizational and innovational renewals,  which bring many companies to the attempt of behaving like startups \cite{Christensen1997}.

The implementation of methodologies to structure and control the development activities in startups is a major challenge for engineers \cite{Coleman2008}. Generally, the management of software development is achieved through the introduction of software processes, which defineswhat steps the development organizations should take at each stage of production and provide assistance in making estimates, developing plans and measuring quality \cite{Chroust1996}. In the last decades, several models have been introduced to control software development activities. However, their application in startup companies doesn't report significant benefits \cite{Coleman2008a, Coleman2008, Sutton2000}.

In this context, software engineering (SE) faces complex and multifaceted obstacles in understanding how to manage development processes. Sutton defines startups as creative and flexible in nature and reluctant to introduce process or bureaucratic measures which may result ineffective \cite{Sutton2000}. Also Bach refers to startups as \textit{``a bunch of energetic and committed people without defined development processes''} \cite{Bach1998}. In fact, startups have very limited resources and typically wish to use them to support product development instead of establishing processes \cite{Coleman2008, Heitlager2007}. Some attempts to taylor lightweight processes to startups, reported basic failure of their application : \textit{``Everyone is busy, and software engineering practices are often one of the first places developers cut corners''} \cite{Martin2007}. Rejecting the notion of repeatable and controlled processes, startups prominently take advantage of unpredictable, reactive and low-precision \cite{surviving-os-cockburn} engineering practices \cite{Sutton2000, Tanabian2005, Chorev2006, Kakati2003}.

Moreover, most startups develop packaged applications rather than software for a specific client \cite{genome2012}. Issues related to this domain are addressed in literature by the area known as \textit{market-driven software development} (sometimes called \textit{packaged software development} or \textit{COTS software development} \cite{Alves2006}). Among other results, researchers emphasize the importance of time-to-market as a key strategic objective \cite{dagMDR,dagMDR,sawyer99} for companies operating in this domain.  Other peculiar aspects which influence the software development are related to requirements, which are reported to be often ``invented by the software company'' \cite{512553}, ``rarely documented'' \cite{Karlsson02challengesin}, and can be validated only after the product is released to market \cite{dahl2003,Keil:1995}. Under these circumstances, failure of product launches are largely due to ``products not meeting customer needs''  \cite{Alves2006}.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Lack of research in startups}
Despite some studies tried to address the above mentioned issues, \cite{SMS} found only a few SE works in this specific area, as confirmed by other studies \cite{Coleman2008, Coleman2008a, Coleman2007, Sutton2000}. Moreover the studies, identified in our systematic review, appear to be highly fragmented and spread across different areas rather than constituting a consistent \textit{body of knowledge}.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Definition of startup}
The word \textit{startup} appeared in the SE literature for the first time in 1994 in an article written by Carmel \cite{Camel1994a} where he studied the \textit{time-to-completion in young package firm}. He noticed how these companies were particularly innovative and successful, advocating for the need of more research investigating software development practices so as to replicate success and try to transfer it to other technology sectors.

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Evolution of the process}

@@

Then, when startups are mature enough to support software process improvement (SPI), the solutions considered according to the state-of-the-art are oriented towards light-weight processes such as a design of development process based on XP, proposed by Zettel in \cite{Zettel2001}. The process consists of a set of activities and artifacts (in addition to some important roles) defined in order to identify responsibilities and tools to utilize. But, despite the promising benefits reported by Zettel, we were not able to identify any future evaluation in real-world settings.

@@

The initial design of the case study is supported by the results of the \textit{systematic mapping study} (SMS) \cite{SMS} which contributed to define the initial questions for both questionnaires and semistructured interviews. The process depicted in Figure \ref{fig:gt:completemethodology} is evolutionary and affects the design at each new iteration.  In \textit{data collection} the empirical results are stored in a single \textit{case study database} and subsequently processed in \textit{data analysis} to form the theoretical categories.  At each iteration the new emergent theory is updated following a formal procedure (\textit{Theory generation}), and after verifying that the \textit{theoretical saturation}\footnote{The point at which executing more interviews wouldn't bring any additional value for constructing the theory.} of categories has been achieved, we finally proceeded \textit{theory validation} or performed another complete cycle.

@@

\item The closest development approach undertaken by early-stage startups is the \textit{Lean startup methodology}.  A highly evolutionary development approach, centered around the quick production of a functioning prototype and guided by customer feedbacks is a central aspect of the \textit{Lean startup methodology} pioneered by Ries \cite{Ries2011}. However, startups do not formally follow the cycle \textit{build-measure-learn} as it is proposed by the methodology, where they should set-up an experiment to \textit{test the riskiest element of the business idea}, nor make explicit use of the good practices suggested in the book. For example, when it comes to user feedbacks, startups do not explicitly follow the step-by-step process of \textit{customer development} formally defined by \cite{Blank2005}. Instead, startups absorb and implement the high-level principles that outlines the methodology, reflected in the model by the theoretical category \textit{find the product/market fit quickly} and its subcategories.

@@

\item We validate the correctness of the GT procedure by attesting that high-level relations between categories, evaluating if they are compliant with the empirical data we collected.

\item Finally we demonstrate how engineering elements independently reported in the follow-up questionnaire by practitioners, can be mapped to theoretical categories in the model.
\end{compactitem}

@@

\item Buy versus build-  the most radical possible solution for constructing software is not to construct it at all, taking advantage of what others have already implemented. It is the main strategy, which enables startups to externalize complexity to third party solutions explained within CAT1,  \textit{speed-up development}.
\item Requirements refinement and rapid prototyping- avoid deciding precisely what to build but rather iteratively extract and refine the product requirements with customers and users. It represents the evolutionary development approach to maintain a fast release-cycles during development (CAT2).
\item Incremental development- starting from simple solutions allows organizations to early prototype and control complexity overtime. It represents the main purpose of focusing on \textit{limited number of suitable functionalities} examined in \textit{product quality has low priority}  category (CAT3).
\item Great team- people are the center of a software project and it is profoundly important to empower and liberate their creative mind. It is the main objective of empowering developers' capabilities described in \textit{team is the catalyst of development}(CAT4).
\end{compactitem}

Brooks' forecast about software development strategies is revealed to be extremely accurate, according to the state-of-practice in modern startups.
%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{High-level relations validity}
\label{sect:theory:validation:stats}

>>>>

\author{FirstName~SecondName,~\IEEEmembership{Fellow,~IEEE,}
        ..~..,~\IEEEmembership{Fellow,~IEEE,}
        and~...~....,~\IEEEmembership{Life~Fellow,~IEEE}% <-this % stops a space
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem Author info.\protect\\
% note need leading \protect in front of \\ to get a newline within \thanks as
% \\ is fragile and will error, could use \hfil\break instead.
E-mail: see http://something/contact.html
\IEEEcompsocthanksitem Author info.}% <-this % stops a space
\thanks{}}

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEcompsoctitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
%\boldmath
Software startups are newly created companies with no operating history and extremely fast in producing cutting-edge technologies. Despite their increasing importance in economy there are only few scientific studies attempting to address software engineering (SE) issues, especially for early-stage startups. This research aims to understand the software development strategies conducted by practitioners, in the period of time that goes from idea conception to the first open beta release. A \textit{grounded theory} approach guided the execution of a case study to explore the state-of-practice. As result, the Greenfield startup model (GSM)  reveals the urgent priority of startups of releasing the product as quickly as possible to verify the product/market fit and to adjust the business and product trajectory according to early collected user feedback. Nevertheless, the need of shortening time-to-market, by speeding-up the development through low-precision and product-centric engineering activities, is counterbalanced by the need of restructuring the product before setting off for further growth.
\end{abstract}
% IEEEtran.cls defaults to using nonbold math in the Abstract.
% This preserves the distinction between vectors and scalars. However,
% if the journal you are submitting to favors bold math in the abstract,
% then you can use LaTeX's standard command \boldmath at the very start
% of the abstract to achieve this. Many IEEE journals frown on math
% in the abstract anyway. In particular, the Computer Society does
% not want either math or citations to appear in the abstract.

@@

Nevertheless, by speeding-up the development process, startups accumulate technical debt, ignoring processes, relying on informal communication and replacing documentation with low-precision artifacts. For startups it is essential to look ahead within short-term deadlines and determine the right product/market fit driven by user feedback, using flexible and reactive development approaches and adopting mitigation strategies to contain the severity of accumulated technical debt.

@@

The remain of this paper is structured as follow:  we provide relevant information about startups in section \ref{backg}. Section \ref{resmet} introduces the research questions and discusses the design and execution of the research process. Results are discussed in section \ref{res:gsm}, which presents the GSM model. Section \ref{res:val}  compares results of this research to literature and frameworks. Section \ref{valt} discusses the validity threats considered during this study. Finally, this paper concludes in section \ref{conc}.

@@

Looking at the number of new business incubators which appeared in the last decade one can evaluate the importance of startups \cite{Grimaldi2005}. The wave of disruption in new technologies has led companies to be more and more competitive, forcing themselves to radical organizational and innovational renewals,  which bring many companies to the attempt of behaving like startups \cite{Christensen1997}.

The implementation of methodologies to structure and control the development activities in startups is a major challenge for engineers \cite{Coleman2008}. Generally, the management of software development is achieved through the introduction of software processes, which define what steps the development organizations should take at each stage of production and provide assistance in making estimates, developing plans and measuring quality \cite{Chroust1996}. In the last decades, several models have been introduced to control software development activities. However, their application in startup companies doesn't report significant benefits \cite{Coleman2008a, Coleman2008, Sutton2000}.

In this context, software engineering (SE) faces complex and multifaceted obstacles in understanding how to manage development processes. Sutton defines startups as creative and flexible in nature and reluctant to introduce process or bureaucratic measures which may result ineffective \cite{Sutton2000}. Also Bach refers to startups as \textit{``a bunch of energetic and committed people without defined development processes''} \cite{Bach1998}. In fact, startups have very limited resources and typically wish to use them to support product development instead of establishing processes \cite{Coleman2008, Heitlager2007}. Some attempts to taylor lightweight processes to startups reported basic failure of their application: \textit{``Everyone is busy, and software engineering practices are often one of the first places developers cut corners''} \cite{Martin2007}. Rejecting the notion of repeatable and controlled processes, startups prominently take advantage of unpredictable, reactive and low-precision \cite{surviving-os-cockburn} engineering practices \cite{Sutton2000, Tanabian2005, Chorev2006, Kakati2003}.

Moreover, most startups develop packaged applications rather than software for a specific client \cite{genome2012}. Issues related to this domain are addressed in literature by the area known as \textit{market-driven software development} (sometimes called \textit{packaged software development} or \textit{COTS software development} \cite{Alves2006}). Among other results, researchers emphasize the importance of time-to-market as a key strategic objective \cite{dagMDR,sawyer99} for companies operating in this domain.  Other peculiar aspects which influence the software development are related to requirements, which are reported to be often ``invented by the software company'' \cite{512553}, ``rarely documented'' \cite{Karlsson02challengesin}, and can be validated only after the product is released to market \cite{dahl2003,Keil:1995}. Under these circumstances, failure of product launches are largely due to ``products not meeting customer needs''  \cite{Alves2006}.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{General lack of research in startups}
Despite some studies tried to address the above mentioned issues, a SMS \cite{SMS} revealed only a few SE works in this specific area, as confirmed by other studies \cite{Coleman2008, Coleman2008a, Coleman2007, Sutton2000}. Moreover, the studies, identified in \cite{SMS}, appear to be highly fragmented and spread across different areas rather than constituting a consistent \textit{body of knowledge}.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Software development in startups}

The word \textit{startup} appeared in the SE literature for the first time in 1994 in an article written by Carmel \cite{Camel1994a} where he studied the \textit{time-to-completion in young package firm}. He noticed how these companies were particularly innovative and successful, advocating for the need of more research investigating software development practices so as to replicate success and try to transfer it to other technology sectors.

@@

%--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------%
\subsection{Software process improvement in startups}

When startups are mature enough to support software process improvement (SPI), the solutions considered according to the state-of-the-art are oriented towards light-weight processes such as a design of development process based on XP, proposed by Zettel in \cite{Zettel2001}. The process consists of a set of activities and artifacts (in addition to some important roles) defined in order to identify responsibilities and tools to utilize. But, despite the promising benefits reported by Zettel, we were not able to identify any future evaluation in real-world settings.

@@

The initial design of the case study is supported by the results of the \textit{systematic mapping study} (SMS) \cite{SMS} which contributed to define the initial questions for both questionnaires and semistructured interviews. The process depicted in Figure \ref{fig:gt:completemethodology} is evolutionary and affects the design at each new iteration.  In \textit{data collection} the empirical results are integrated and stored in a single \textit{case study database} and subsequently processed in \textit{data analysis} to form the theoretical categories.  At each iteration the new emergent theory is updated following a formal procedure (\textit{Theory generation}), and after verifying that the \textit{theoretical saturation}\footnote{The point at which executing more interviews wouldn't bring any additional value for constructing the theory.} of categories has been achieved, we finally proceeded to the \textit{theory validation} or performed another complete cycle.

@@

\item The closest development approach undertaken by early-stage startups is the \textit{Lean startup methodology}.  A highly evolutionary development approach, centered around the quick production of a functioning prototype and guided by customer feedback is a central aspect of the \textit{Lean startup methodology} pioneered by Ries \cite{Ries2011}. However, startups do not formally follow the cycle \textit{build-measure-learn} as it is proposed by the methodology, where they should set-up an experiment to \textit{test the riskiest element of the business idea}, nor make explicit use of the good practices suggested in the book. For example, when it comes to user feedback, startups do not explicitly follow the step-by-step process of \textit{customer development} formally defined by \cite{Blank2005}. Instead, startups absorb and implement the high-level principles that outlines the methodology, reflected in the model by the theoretical category \textit{find the product/market fit quickly} and its subcategories.

@@

\item We validate the correctness of the GT procedure by attesting the relations with the GSM model's core category, evaluating if they are compliant with the empirical data we collected.

@@

\item Buy versus build-  the most radical possible solution for constructing software is not to construct it at all, taking advantage of what others have already implemented. It is the main strategy, which enables startups to externalize complexity to third party solutions explained within CAT1,  \textit{speed-up development}.
\item Requirements refinement and rapid prototyping- avoid deciding precisely what to build but rather iteratively extract and refine the product requirements with customers and users. It represents the evolutionary development approach to maintain a fast release-cycles during development (CAT2).
\item Incremental development- starting from simple solutions allows organizations to early prototype and control complexity overtime. It represents the main purpose of focusing on \textit{limited number of suitable functionalities} examined in \textit{product quality has low priority}  category (CAT3).
\item Great team- people are the center of a software project and it is profoundly important to empower and liberate their creative mind. It is the main objective of empowering developers' capabilities described in \textit{team is the catalyst of development} (CAT4).
\end{compactitem}

Brooks' forecast about software development strategies is revealed to be extremely accurate, according to the state-of-practice in modern startups.
%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{Theoretical categories and existing literature}
\label{sect:theory:validation:sms}

In this section we report the evaluation of our theory in contrast with main relevant contribution of studies retrieved within \cite{SMS}, discussing the categories of the GSM model.

The comparison has been executed by identifying issues related to the theoretical categories in the retrieved relevant studies. We illustrate how the selected studies contributed to our framework validation as following: for each retrieved article we present the main results, mapping them to categories in accordance to the theoretical framework. Table \ref{tab:an:literature_comp} shows the main contributions of the 37 identified studies, where an 'X' is placed in correspondence to the theoretical category they considered. The table has been sorted by computing how many categories of our theoretical frameworks have been considered by the study (column \textit{Count}). Therefore in the highest part of the table there are the studies which somehow discuss the higher number of concepts identified in the early-stage startups framework.

\begin{table*}[!t]
\renewcommand{\arraystretch}{1.3}
\caption{SMS \cite{SMS} and GSM} 
\label{tab:an:literature_comp}
\centering
\begin{tabular}{|l||c||c||c||c||c||c||c||c||c|}

    \hline
    Author (year) & CAT1  & CAT2  & CAT3  & CAT4  & CAT5  & CAT6  & CAT7  & Count & Ref. \\
\hline
Sutton (2000) & X     & X     & X     & X     & X     & X     & X     & 7     & \cite{Sutton2000} \\
Kajko (2008) & X     & X     & X     & X     & X     & X     & X     & 7     & \cite{Kajko-Mattsson2008} \\
Crowne (2002) & X     & X     & X     & X     & X     & X     & X     & 7     & \cite{Crowne2002} \\
Coleman (2008) & X     & X     & X     & X     & X     & X     & X     & 7     & \cite{Coleman2008a} \\
Coleman (2008) & X     & X     & X     & X     & X     & X     & X     & 7     & \cite{Coleman2008} \\
Coleman (2007) & X     & X     & X     & X     & X     & X     & X     & 7     & \cite{Coleman2007} \\
Camel (1994) & X     & X     & X     & X     & X     & X     & X     & 7     & \cite{Camel1994a} \\
Yoffie (1999) & X     & X     & X     & X     & X     & X     &       & 6     & \cite{Yoffie1999} \\
Zettel (2001) & X     & X     & X     & X     &       &       & X     & 5     & \cite{Zettel2001} \\
Jansen (2008) & X     &       & X     &       & X     & X     & X     & 5     & \cite{Jansen2008} \\
Heitlager (2007) &       &       & X     & X     & X     & X     & X     & 5     & \cite{Heitlager2007} \\
Deias  (2002) & X     & X     &       & X     & X     & X     &       & 5     & \cite{Deias} \\
Ambler (2002) & X     & X     &       &       & X     & X     & X     & 5     & \cite{Ambler2002} \\
Wood (2005) & X     &       & X     & X     &       &       & X     & 4     & \cite{Wood2005} \\
Tingling (2007) &       & X     &       & X     & X     & X     &       & 4     & \cite{Tingling2007} \\
Taipale (2010) & X     & X     &       & X     & X     &       &       & 4     & \cite{Taipale2010} \\
Silva (2005) & X     & X     &       & X     & X     &       &       & 4     & \cite{Silva2005} \\
Mirel (2000) & X     &       & X     & X     & X     &       &       & 4     & \cite{Mirel2000} \\
Midler (2008) & X     & X     &       & X     &       &       & X     & 4     & \cite{Midler2008} \\
Tanabian (2005) & X     &       &       & X     &       &       & X     & 3     & \cite{Tanabian2005} \\
Stanfill (2007) & X     &       &       &       & X     &       & X     & 3     & \cite{Stanfill2007} \\
Mater (2000) & X     & X     &       &       &       & X     &       & 3     & \cite{Mater2000} \\
Kuvinka (2011) & X     & X     &       & X     &       &       &       & 3     & \cite{Kuvinka2011} \\
Deakins (2005) & X     & X     &       & X     &       &       &       & 3     & \cite{Deakins2005} \\
Yogendra (2002) & X     & X     &       &       &       &       &       & 2     & \cite{Yogendra2002} \\
Wall (2001) & X     &       &       &       &       &       & X     & 2     & \cite{Wall2001} \\
Su-Chuang (2007) & X     &       &       & X     &       &       &       & 2     & \cite{Su-Chan2007} \\
Steenhuis (2008) &       &       &       & X     &       &       & X     & 2     & \cite{Steenhuis2008} \\
Sau-ling Lai (2010) &       & X     &       &       &       &       & X     & 2     & \cite{Lai2010} \\
Kakati (2003) &       & X     &       & X     &       &       &       & 2     & \cite{Kakati2003} \\
Himola (2003) & X     &       &       &       &       &       & X     & 2     & \cite{Hilmola2003} \\
H\"{a}sel (2010) & X     &       &       & X     &       &       &       & 2     & \cite{Hasel2010} \\
Hanna (2010) & X     &       &       &       &       &       & X     & 2     & \cite{Hanna2010} \\
Bean (2005) & X     &       &       &       & X     &       &       & 2     & \cite{Bean2005} \\
Kim (2005) &       & X     &       &       &       &       &       & 1     & \cite{Kim2005} \\
Fayad (1997) &       &       &       & X     &       &       &       & 1     & \cite{Fayad1997} \\
Chorev (2006) &       &       &       & X     &       &       &       & 1     & \cite{Chorev2006} \\

According to our framework, only 7 studies, out of 37 (18.91\%), discuss aspects related to the overall software development process in early stage startups. The remaining part of the studies is only partially covering the phenomenon described by the framework. Looking at the results, we were able to extrapolate information which confirmed the GSM categories.

We can see how the majority of the retrieved studies (29) mention issues related to \textit{speed-up development} which confirms, in fact, the importance of the \textit{core} category. On the other hand we can observe that only a limited number of studies mentions results related to CAT3, CAT5 and CAT6. This suggests the direction of possible future primary studies towards: the technical debt and its future consequences on performance; aspects related to functionality and quality of the first version of the product.

These results confirm the relevance of development teams as widely discussed in the SE literature, here mapping 26 studies to CAT4 (\textit{team catalyst for development}). The importance of people has been widely discussed in other studies in SE advocating for the need of  \textit{empowering people}, which is the element with the greatest impact on software development. Furthermore, being humans \textit{non-linear variables}, they are unlikely to follow repeatable prescriptive methodologies. As stated by Highsmith, \textit{``A drawback of each and every methodology is to expect people to behave consistently over time, when is clearly not like that.''} \cite{Highsmith2000}. For more detailed discussion about the importance of people the reader is referred to some prominent authors: Cooper \cite{Cooper1986}, DeMarco \cite{peopleware-demarco1999}, Coleman \cite{Coleman2004}, Valtanen \cite{Valtanen2008},  Adolph \cite{Adolph2011} and Cockburn \cite{people-first-order}. Especially in early-stage startups, where the whole company overlaps with the development team, factors related to people are without any doubt the most important ones (confirmed with the empirical data of  our case study.

In addition to the theoretical framework proposed by Coleman previously discussed, in the next part of this subsection we present similarities and contrasts with other relevant studies selected with the SMS.

Starting from the concept of \textit{severe lack of resources} many authors describe startups facing with inescapable constraints defined by engineering/business concerns and constant time-pressure \cite{Camel1994a, Cugola98softwareprocesses:}. Because of their young and inexperienced working conditions, startups present limited resources in terms of management and strategic alliances \cite{Sutton2000}.

Crowne in \cite{Crowne2002} identifies four main stages of a startup's lifecycle: startup, stabilization, growth and maturity. Inexperienced practitioners operating in a chaotic environment characterizes the first phase, where there is no strategic plan for developing the product. But, moving towards the \textit{stabilization} phase, the product starts to be unreliable and requirements unmanageable. Consequently, when startups move to the \textit{growth} phase, they initiate structuring and controlling processes, which are incrementally integrated in the development environment. The \textit{maturity} represents the moment when the company is ready to introduce process improvement initiatives. This study profoundly corresponds to different conceptualization of the theoretical framework. Starting from the minimal and low-precision engineering elements, startups start to grow the accumulated technical debt which requires fast integration of more structured and standard processes overtime.
Issues related to low-precision engineering practices during companies' growth are discussed in an experience paper by Ambler \cite{Ambler2002} where he reports the case studies of two growing internet startups approaching an IPO. Despite the two companies were facing business and team growth, attitudes against the introduction of processes to structure and model the system were evident. The need to re-architect and redevelop the codebase lead the researchers to suggest a tailored version of the RUP development framework. Nevertheless, even with the aid of CASE tools\footnote{CASE tools are a class of software that automates many of the activities involved in various engineering activities. See \url{http://www.unl.csi.cuny.edu/faqs/software-enginering/tools.html} .}, startups failed in practicing engineering activities in view of eccesive lack of flexibility. In fact, developers fast moved to simpler tools such as the whiteboard, where they were able to provide input and insights, question assumptions, share past experiences, and even talk about potential implementation strategies. In general the development remained highly iterative, where project teams started to have a little modeling and testing only when necessary. They didn't take the serial approach of completely developing and accepting the requirements model, analysis model and so on. This let them react swiftly and adapt rapidly according to changes into their highly competitive markets. The researchers report \textit{``modeling was streamlined''}, that is modeling what was needed and go straightforward to coding, avoiding analysis paralysis.

The use of whiteboards in working environment is in support of the \textit{use of well-integrated and simple tools}. In both the startups studied by Ambler, documentation was minimal, finding the ``sweet spot'' to have just barely enough tools and artifacts to meet their needs and no more. Generally, the use of \textit{keep a simple and informal workflow} endorses the use of low-precision engineering practices which lead to speed-up development initially, but with negative effects on the accumulated technical debt of a minimal project management and consequently on the final company growth.

Kajko-Mattsson et al. investigated a Swedish software startup company involved in mobile applications \cite{Kajko-Mattsson2008}, reporting the following issues: a lack of requirements gathering process; release cycles and their length were not defined; the releases and their scope were not planned; lack of control over the change requests; lack of process control in terms of  absence of any documentation to track the status and progress of the process; defective releases in terms of lack of testing; poor communication since the communication inside the company was informal and not documented at all. In the same study they designed a process improvement model, based on three main phases: \textit{Evaluate}, \textit{Plan}  and \textit{Change}.

During the evaluation phase Kajko-Mattsson et al. tried to establish the development workflow status within the company, and the authors document a status of chaos where any initiatives of quantitative methodology assessment would have been a waste of time. Therefore, they applied a qualitative methodology approach to define a model of the release management process. In fact, they moved towards the plan phase where the researchers decided to develop a model to manage releases, introducing a process composed by: release scope preparation, release planning, release development, system testing, acceptance testing and release deployment. Despite the effort spent from the researcher to establish a development workflow, practices were not fully conducted even though some marginal benefits have been reported: planned major release from four to six months; planned minor releases from two to four weeks; encompassing urgent corrective releases from one to five days. Regarding the previously mentioned problems, only partial improvements have been achieved since the major obstacle was to motivate employees to change their habits. With perfect hindsight the researchers claim that understanding and adapting solutions by solving one problem at a time would have increased benefits. Then, despite the reported results, start establishing simple and informal workflow aided the organization to move from the stage of knowing nothing to knowing at least something about their development process.

The integration of minimal project management during the development activities can foster the control over development productivity and control the \textit{accumulated technical debt}, but still with particular attention to the development workflow status within the company, designing specific and tailored solutions.

In fact, prescribed development models such as SPI in startups are basically ignored by startups, as discussed in \cite{Zettel2001}. To overcome the need of structure and control to accelerate the development process, in his study, Zettel developed a \textit{``LIghtweight Process for E-business software development''} (LIPE). This lightweight development method integrates XP approach with ideas from the areas of software measurement for project control and process improvement as a compromise between ad-hoc and more rigorous approach. LIPE, as argued from the author, provides scalability of the development process with a certain degree of flexibility, omitting parts of prescribed practices. Nevertheless, we couldn't retrieve any empirical evaluation of the study.

In \cite{Deakins2005}, the author designed a new spiral methodology evolved during fieldwork with a developer team that had limited resources managing innovative product in volatile e-commerce environments. He proposes fast evolutionary approach experimentation with flexible processes obtaining customer's feedback as soon as possible for continuous product improvement. Even in this case the model is compliant with the nature of a random, ad-hoc and visionary product development. In fact, evolutionary development approaches are typically more suitable  for web applications, as confirmed in \cite{Deshpande2001}. Then the practices are oriented towards rapid and iterative design with complete freedom to the developer team, enabled to operate anywhere within the organization, according also to the category \textit{team is the catalyst of development}.

Carmel reports in \cite{Camel1994a} the need of flexible and rapid development solutions to shorten time-to-market, but even more important, the need of high-skilled team developers, which heavily impact on \textit{speeding-up development}. In fact, the author suggests that entrepreneurs need to look for a well formed, skilled core development team and not just a set of product ideas and features. Moreover the team must be empowered with full-stack and self-organization settings with the flexibility of minimal bureaucracy. A studied company stated: \textit{``we didn't need weeks or months of detailed modeling and documentation , but rather modeling the architecture a little and then either exploring strategies by providing users with code or simply starting work on the actual software itself''}.

Eventually to focus on simple solutions, two other studies \cite{Wall2001, Bean2005} sugges to \textit{``reuse the wheel, instead of reinventing it''}, taking advantage of open source software whenever it is possible, building systems quickly and effectively as a result. The author conclude stating that even though the principles in leading-edge development processes are the same principles of yesteryear, the way in which the fundamentals are applied has changed from prescriptive to agile solutions with less concern to quality aspects, also confirmed in \cite{Mater2000}.

Thus, by summarizing the comparison, we have found that most of the contributions of the studies are compliant with the behavioural model of the phenomenon we provided.

%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%%--------------------------%
\subsection{Confounding factors from the literature}
\label{sect:an:comp:cat-vs-literature:conf}

The purpose of this subsection is to identify which confounding factors need to be taken into consideration when evaluating the theoretical framework. The confounding factors are those variables which and have not been considered in the model and might interfere with the theoretical framework positively or negatively \cite{ColinRobson2009}. Even though there are no systematic procedure to identify confounding factors, we report those that have been explicitly identified by the literature review, that are related to: \textit{creativity, innovation, market type and application type}.

A prominent contribution investigating creativity and innovation in startups, is presented in \cite{Heitlager2007}. The study reports how development approaches oriented to the product in the early-stage life-cycle, called \textit{fluid} (see Figure \ref{fig:an:comp:cof:inn}), have main impact to innovative solutions initially. After a while the need of structuring and controlling the growing company size (along with \textit{transition} and \textit{specific} phases) arises the need of project management activities and software processes for long-term scalability issues.

Based on the model presented in \cite{Heitlager2007}, it identifies the lifecycle stages of a startup (\textit{fluid}, \textit{transition} and \textit{specific}), corresponding for each stage a development approach (product-oriented and process-oriented) the level of innovation achieved. The study reports how product-oriented development, in contrast with forced and mature process imposition, gives degree of freedom to the development team, enhancing \textit{creativity} of developers and augmenting the \textit{innovation} capability of the company in the early-stage.

The empirical model we presented has been obtained from the transcripts of interviews, where the theme of innovation only seldom emerged. However, if innovative and creative factors were considered in the model, the integration of results would have been straightforward, since the evolution of this model is totally compliant with our assumption of CAT5 and CAT6. From product-oriented development in the \textit{fluid} phase to the process-oriented development in the \textit{specific} phase, the \textit{innovation} confounding factor seems to behave initially the same as the \textit{speed-up development}. Nevertheless, we were unable to ground these evidences in the GT data even though they might present strict correlation with the choice of an early product-development orientation with low-precision engineering elements.

Other two important factors, that are not grounded in our empirical data, are related to: \textit{a.} the requirements expected by a specific market sector (also called \textit{market requirements} in \cite{Coleman2007}); and  \textit{b.} the application type as reported in \cite{Sutton2000, Coleman2008, Coleman2008a}. The main impact of those confounding factors is related to the adoption of flexible and reactive solutions for the development process. In particular they refer to the necessity of fulfillment of quality concerns that goes beyond scalability and UX defined by the framework. Especially when requirements are rigidly imposed or the application domain is well-known, providing low-quality products to final users might determine the failure of a startup.

For example security in critical applications such as safety-critical systems present quality concerns, which must be granted. But also minor risky application domains such as e-commerce systems have high-impact factors for obtaining the first customer reference, as discussed in \cite{Deakins2005,Kim2005,Silva2005}. This characterizes those applications, whose part of functionalities are already successfully implemented in other software systems or whose failure might report serious financing or social damages \cite{Bass2003}.Yet, limiting our context to early-stage software startups, which typically develop first cutting-edge product in beta versions, partially mitigates the high-quality demands of users.

To understand how to obtain processes in accordance to a set of customer needs, correlating the proper levels of customers' satisfaction, we introduce the Kano model, described in \cite{Ullah2011} and extensively studied in product development.  As shown in the Kano model, users preferences can be classified into four categories: \textit{basic}, \textit{excitement}, \textit{performance} and \textit{indifference} areas.
